---
title: "Week 14 Exercises: Quarto document"
pagetitle: "Week 14 Exercises"
date: 2026-11-23
date-modified: last-modified
author:
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://go.osu.edu/bioinformatics-core
      - name: Dept. of Plant Pathology, Ohio State University
        url: https://plantpath.osu.edu
language:
  title-block-published: "Week starts on"
  title-block-modified: "Last modified"
---

------------------------------------------------------------------------

<br>

## Setting up

- At <https://ondemand.osc.edu>, start a new RStudio session
  ([Starting an RStudio Server session at OSC](../week11/w11a_r-basics.html#starting-an-rstudio-server-session-at-osc)).

- Switch to your RStudio Project
  ([Working directory and RStudio Projects](../week11/w11a_r-basics.html#Working-directory-and-RStudio-Projects))

- Create a new directory `week14` in your home directory using function
  `dir.create("week14")`

- Open a new `Quarto file` in RStudio:
  `File` -> `New File` -> `Quarto Documents` and save the file as
  `w14_exercises.qmd` (inside `week14` directory). You will write your answers
  to the questions below in this Quarto file

## Exercise

### YAML header

1. Save your Quarto (.qmd) file that contains your script in HTML format.

2. Add the following key–value pairs to the YAML header of your Quarto document:

- Author name
- Date
- Table of contents

3. Create nested format options in the YAML header using `format:` `html:` and set the `theme to cyborg`.

### Code Chunks

4. Install the `palmerpenguins` package in the Console, and load the package in a code chunk of your Quarto document. Save the penguins dataset in an object called `penguin_data`.

5. Open the second code chunk, select the columns `flipper_length_mm`, `island` , `year` and save them as `new_penguins` object. Add a code chunk option to hide the code in the rendered file.

6. Use `penguin_data` to create a scatter plot with the `ggplot2` package of `tidyverse`. Set 
`flipper_length_mm` as the **x-axis** and `body_mass_g` as the **y-axis**. Refer to the `ggplot2` 
documentation to understand the syntax of scatter plot. We will discuss `ggplot` in `Week 13`.

7. Customize your figure by:
- Adding a figure caption
- Changing the figure dimensions (height and width)
- Aligning the figure to the center using chunk options

8. Render your document, download the output, and verify that it includes the scatter plot.

9. If document does not contain scatter plot, add the appropriate option in the YAML header to include figures in the HTML output.

10. Render the document again and confirm that the scatter plot is now included.

### Text

11. Add headers of different levels above your code chunks.

12. Write a short text section describing the scatter plot you created in question 9.

## Solutions

### YAML header

<details><summary>**1-3.** YAML header</summary>

```{r eval=FALSE}
##The solution of these questions are provided in the exercise of week14 lecture.
```
</details>

### Code chunks

<details><summary>**4.** install `palmerpenguins` package and save object </summary>

```{r}
library(palmerpenguins)
penguin_data <- penguins
```
</details>

<details><summary>**5.** Select columns </summary>
   
```{r}
## First load tidyverse to use select function
library(tidyverse)
new_penguins <- penguin_data |> 
  select( flipper_length_mm, island , year)
```
</details>


<details><summary>**6.** Scatter plot </summary>

```{r}
library(ggplot2)
penguin_data |> 
  ggplot(aes(flipper_length_mm, body_mass_g)) +
  geom_point()
```
</details>

<details><summary>**7.** Customize your plot </summary>

```{r}
#| fig-cap: "Scatter plot of palmerpenguins datasets"
#| fig-width: 5
#| fig-height: 4
#| fig-align: center
#| echo: fenced
penguin_data |> 
  ggplot(aes(flipper_length_mm, body_mass_g)) +
  geom_point()
```
</details>

<details><summary>**8.** Include your plot in HTML </summary>

**`embed-resources: true` in your YAML header to make your document self-contained**

</details>

