---
title: "Week 12 Exercises: R Basics"
pagetitle: "Week 12 exercises"
date: 2026-11-09
date-modified: last-modified
author:
  - name: Menuka Bhandari
    email: bhandari.72@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://go.osu.edu/bioinformatics-core
      - name: Dept. of Plant Pathology, Ohio State University
        url: https://plantpath.osu.edu
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
other-links:
  - text: "Week 12 overview"
    icon: check2-all
    href: w12_overview.html
  - text: "w12a: Intro to R and RStudio"
    icon: book
    href: w12a_r-basics.html
  - text: "w12b: R data structures and data types"
    icon: book
    href: w12b_data.html
language:
  title-block-published: "Week starts on"
  title-block-modified: "Page last modified"
  other-links-title: "Also this week"
---

--------------------------------------------------------------------------------

<hr style="height:1pt; visibility:hidden;" />

## Setting up

- At <https://ondemand.osc.edu>, start a new RStudio session
  ([Starting an RStudio Server session at OSC](./w12a_r-basics.qmd#starting-an-rstudio-server-session-at-osc))

- Switch to your `week12` RStudio Project
  ([Working directory and RStudio Projects](./w12a_r-basics.qmd#working-directory-and-rstudio-projects))

- Open a new R script in RStudio:
  `File` -> `New File` -> `R Script` and save the R script as
  `w12_exercises.R` (inside `week12`).
   You will write your answers to the exercises below in this R script.

## Part A: Create variables and perform arithmetic operations

1. Assign the value `5` to the variable `week11`, and `10` to the variable `week12`.
   Print these variables to the console.

2. Add the values of `week11` and `week12` and assign the result to a new variable `total_weeks`.
   Similarly, multiply the values of `week11` and `week12` and assign the result
   to a new variable `multiplied_weeks`.
   Print these variables to the console.

## Part B: Create vectors and change their elements

3. Create a vector `v1` that contains the numbers 1 through 5.
   Then, extract the third element of `v1`, printing it to the console.
   Likewise, create another variable `v2` with numbers 6 to 10.
   Check what data types `v1` and `v2` contain.

4. Change the value of the third element of `v1` to "Midweek" and print `v1`.
   Check the data type of `v1` after changing the third element to `Midweek`.
   Try to change the `v1` to `numeric` and see what happens.

5. Create a vector `v3` that combines `v1` and `v2`.
   Extract the third, fourth and fifth element of `v3` and assign it to `v4`.
   Print `v4` to the console.

## Part C: Install and load new packages in R

6. Install the package `palmerpenguins` and load the package in R.
   We will use this package in `week13` for data visualization.

## Solutions

### Part A: Create variables and perform arithmetic operations

<details><summary>**1.** Assign values to variables</summary>

```{r}
week11 <- 5
week11

week12 <- 10
week12
```
</details>

<details><summary>**2.** Arithmetic operations</summary>

```{r}
total_weeks <- week11 + week12
total_weeks

multiplied_weeks <- week11 * week12
multiplied_weeks
```
</details>

### Part B: Create vectors and change their elements

<details><summary>**3.** Create vectors and extract elements</summary>

```{r}
v1 <- 1:5
v1[3]

v2 <- 6:10

str(v1)
str(v2)
```
</details>

<details><summary>**4.** Change vectors </summary>

**After replacing a numeric element with a string, the entire vector becomes a character vector**
**And when you try to convert it back to numeric, R will return `NA` for the non-numeric elements:**
   
```{r}
v1[3] <- "Midweek"
v1
str(v1)
as.numeric(v1)
```
</details>

<details><summary>**5.** Combine vectors </summary>

```{r}
v3 <- c(v1, v2)
v4 <- v3[3:5]
v4
```
</details>

### Part C: Install and load new packages in R

<details><summary>**6.** Install and load a package </summary>

```{r, eval=FALSE}
install.packages("palmerpenguins")
library(palmerpenguins)
```
</details>

