---
title: "R vectors and data types"
subtitle: "Week 12 -- Lecture B"
pagetitle: "w12b: R data"
date: 2026-11-09
date-modified: last-modified
author:
  - name: Menuka Bhandari
    #orcid: TBA
    email: bhandari.72@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://go.osu.edu/bioinformatics-core
      - name: Dept. of Plant Pathology, Ohio State University
        url: https://plantpath.osu.edu
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
other-links:
  - text: "Week 12 overview"
    icon: check2-all
    href: w12_overview.html
  - text: "w12a: Intro to R and RStudio"
    icon: book
    href: w12a_r-basics.html
  - text: "Week 12 exercises"
    icon: pencil
    href: w12_exercises.html
language:
  title-block-published: "Presented on"
  title-block-modified: "Page last modified"
  other-links-title: "Also this week"
editor: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

<br>

## Introduction

### Overview & learning goals

In this lecture, you will learn about some of the nuts and bolts of R,
which will give you a solid foundation to start doing more exciting
things like data wrangling, analysis, and visualization.

Specifically, you will learn:

-   Several kinds of **data structures** - vectors (*this lecture*) and
    data frames and matrices (*next lecture*)
-   R's main **data types** - `character`, `integer`, `double` and
    `logical`

### R's data structures and data types

-   **Data structures** are the kinds of objects that R can store data
    in. Here, you'll learn about one of the most common ones, the
    *vector*. Next week, you'll learn about another very common data
    structure: the *data frame*, and a less common one, the *matrix*.

-   **Data types** are how R distinguishes between different kinds of
    data like numbers versus character strings. Here, we'll talk about
    the 4 main data types: `character`, `integer`, `double`, and
    `logical`.

An analogy: If this was about food rather than data, you can think of
different data types as different food items, and various data
structures as different types of containers to store food in.

### Setting up

1.  At <https://ondemand.osc.edu>, start an RStudio session like in the
    previous lecture ([Starting an RStudio Server session at
    OSC](./w12a_r-basics.html#Working-directory-and-RStudio-Projects))
2.  Switch to your `week12` RStudio Project created in previous lecture
    ([Working directory and RStudio
    Projects](./w12a_r-basics.html#Working-directory-and-RStudio-Projects))
3.  Open a new R Script and save it as `week12b.R` in your `week12`
    folder ([R scripts](./w12a_r-basics.html#R-scripts))

## Vectors

The R data structure we will explore in this lecture is the simplest
one: the vector. A vector in R is essentially **a collection of one or
more items**. Moving forward, we'll call such individual items
"elements".

### Single-element vectors (and quoting)

Vectors can consist of just a single element, so each of the two lines
of code below creates a vector:

```{r}
vector1 <- 8
vector2 <- "panda"
```

In the `"panda"` example, which is a **character string** (*string* for
short):

-   `"panda"` constitutes *one element*, not 5 (its number of letters).
-   Unlike when dealing with numbers, we have to *quote the string*.[^1]

[^1]:  Either double quotes (`"..."`) or single quotes (`'...'`) work,
    but the former are most commonly used by convention.

Character strings need to be quoted because they are otherwise
interpreted as R objects -- for example, because our vectors `vector1`
and `vector2` are objects, we refer to them without quotes:

```{r}
# [Note that R will show auto-complete options after you type 3 characters]
vector1
vector2
```

Meanwhile, the code below doesn't work because there is no *object
called `panda`*:

```{r, error=TRUE}
vector_fail <- panda
```

### Multi-element vectors

A common way to make vectors with **multiple elements** is by using the
`c` (combine) function:

```{r}
c(2, 6, 3)
```

::: {.callout-note appearance="simple"}
*Unlike in the first couple of vector examples,* *we didn't save the
above vector to an object:* *now the vector simply printed to the
console (similar to the standard output in shell) -- but it is created
all the same.*
:::

`c()` can also **append** elements to an existing vector:

```{r}
# First we create a vector:
vector_to_append <- c("cardinal", "chickadee")
vector_to_append

# Then we append another element to it:
c(vector_to_append, "bald eagle")
```

To create vectors with **series of numbers**, a couple of shortcuts are
available. First, you can make series of whole numbers (integers) with
the `:` operator:

```{r}
1:10
```

Second, you can use a function like `seq()` and its arguments `from`
(starting value), `to` (end value), and `by` (step size) for fine
control over the sequence:

```{r}
stepwise_vec <- seq(from = 6, to = 8, by = 0.2)
stepwise_vec
```

### Vectorization

Consider the output of this command:

```{r}
stepwise_vec * 2
```

Above, **every individual element in `stepwise_vec` was multiplied by
2**. We call this behavior "vectorization" and this is a key feature of
the R language. (Alternatively, you may have expected this code to
*repeat* `stepwise_vec` twice, but this did not happen!)

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} Exercise: Vectors

Make a vector `x` with the whole numbers 1 through 26. Then, subtract
0.5 from each element in `x` and save the result in vector `y`. Check
your results by printing both vectors.

<details>

<summary>Click for the solution</summary>

```{r}
x <- 1:26
x

y <- x - 0.5
y
```

</details>

<hr style="height:1pt; visibility:hidden;" />

------------------------------------------------------------------------

*Bonus:* What do you think the result of the following operation will
be? We didn't cover this kind of scenario, but go ahead and test your
intuition! After you've decided on your expectation, run the code and
check if you were correct.

```{r, eval=FALSE}
1:5 * 1:5
```

<details>

<summary>Click for the solution</summary>

```{r}
1:5 * 1:5
```

Both vectors are of length 5 which will lead to "element-wise matching":
the first element in the first vector will be multiplied with the first
element in the second vector, the second element in the first vector
will be multiplied with the second element in the second vector, and so
on.

</details>
:::

### Exploring vectors

R has many functions that provide information about vectors and other
types of objects, such as:

-   Get the **number of elements** with `length()`:

    ```{r}
    length(stepwise_vec)
    ```

-   See the **first and last few elements**, respectively, with `head()`
    and `tail()`:

    ```{r}
    # Print the first 6 elements:
    head(stepwise_vec)

    # Print the last 6 elements:
    tail(stepwise_vec)

    # Both head and tail have argument `n` to specify the number of elements:
    tail(stepwise_vec, n = 2)
    ```

-   Get **arithmetic summaries** like `mean()` for vectors with numbers:

    ```{r}
    # mean() will compute the mean (average) across all elements
    mean(stepwise_vec)
    ```

### Extracting elements from vectors

Extracting element from objects like vectors is often called
**"indexing"**. In R, we can do this using "bracket notation" with
square brackets <kbd>\[ \]</kbd> -- for example:

-   Get the second element with `[2]`[^2]:

    ```{r}
    stepwise_vec[2]
    ```

-   Get the second through the fifth elements with `[2:5]`:

    ```{r}
    stepwise_vec[2:5]
    ```

-   Get the first and eight elements with `[c(1, 8)]`:

    ```{r}
    stepwise_vec[c(1, 8)]
    ```

[^2]:  R uses 1-based indexing, which means it starts counting at 1 like
    humans do. Index 2 therefore simply corresponds to the second
    element. Python and several other languages use 0-based indexing,
    which starts counting at 0 such that the second element corresponds
    to index 1.

::: {.callout-tip appearance="simple"}
To put this in a generalized way: we can extract elements from a vector
by using another vector, whose values are the positional indices of the
elements in the original vector.
:::

::: callout-note
### Changing vector elements using indexing

Above, we saw how you can *extract* elements of a vector using indexing.
To *change* elements in a vector, simply use the bracket on the other
side of the arrow -- for example:

-   Change the first element to `30`:

    ```{r}
    stepwise_vec[1] <- 30
    stepwise_vec
    ```

-   Change the last element to `0`:

    ```{r}
    stepwise_vec[length(stepwise_vec)] <- 0
    stepwise_vec
    ```

-   Change the second element to the mean value of the vector:

    ```{r}
    stepwise_vec[2] <- mean(stepwise_vec)
    stepwise_vec
    ```
:::

## Data types

### R's main data types

R distinguishes between different kinds of data, such as character
strings and numbers, using several pre-defined "data types". R's
behavior in various operations depends heavily on the data type -- for
example, the below fails:

```{r, error=TRUE}
"valerion" * 5
```

We can ask what type of data something is in R using the `typeof()`
function:

```{r}
typeof("valerion")
```

R set the data type of `"valerion"` to `character`, i.e. a (character)
string. The earlier command failed because R can't perform mathematical
functions ("*binary operator*") on vectors of type `character`
("*non-numeric argument*").

The `character` data type most commonly contains letters, but anything
that is placed between quotes (`"..."`) will be interpreted as this data
type -- even plain numbers:

```{r}
typeof("5")
```

Besides `character`, three other **common data types** are:

-   `double` / `numeric` -- numbers that can have decimal points:

    ```{r}
    typeof(3.14)
    ```

-   `integer` -- whole numbers only:

    ```{r}
    typeof(1:3)
    ```

-   `logical` (either `TRUE` or `FALSE` -- unquoted!):

    ```{r}
    typeof(TRUE)
    ```

Here is an overview in table format:

| Data type            | Abbreviation  | Explanation                          |
|-------------------|-----------------|-------------------------------------|
| `double` / `numeric` | `dbl` / `num` | Numbers that can have decimal points |
| `integer`            | `int`         | Whole numbers                        |
| `character`          | `chr`         | Character strings                    |
| `logical`            | `lgl`         | `TRUE` or `FALSE`                    |

: {.striped .hover}

<hr style="height:1pt; visibility:hidden;" />

::: callout-note
#### Logicals can be represented as numbers

Consider the following R behavior:

```{r}
TRUE + TRUE
```

```{r}
FALSE + FALSE
```

```{r}
FALSE + 12
```

So, logicals can be used as if they were numbers, where `FALSE`
represents 0 and `TRUE` represents 1.
:::

### A vector can only contain one data type

A vector can only **be composed of a single data type**. As we saw
above, R silently picks the "best-fitting" data type when you create a
vector.

::: exercise
#### {{< fa user-edit >}} Exercise: Data types

In each line below, what do you think the data type (if any) will be?
Try it out and see if you were right.

```{r, eval=FALSE}
typeof("TRUE")
typeof(banana)
typeof(c(2, 6, "3"))
```

<details>

<summary>Click for the solutions</summary>

<hr style="height:1pt; visibility:hidden;" />

1.  `"TRUE"` is `character` (and not `logical`) because of the quotes
    around it:

    ```{r}
    typeof("TRUE")
    ```

------------------------------------------------------------------------

2.  Recall the earlier example: this returns an error because the object
    `banana` does not exist. Any unquoted string (that is not a special
    keyword like `TRUE` and `FALSE`) is interpreted as a reference to an
    object in R.

    ```{r, error=TRUE}
    typeof(banana)
    ```

------------------------------------------------------------------------

3.  This produces a `character` vector, and we'll talk about why in the
    next section:

    ```{r}
    typeof(c(2, 6, "3"))
    ```

</details>
:::

### Type coercion and conversion

R's behavior of returning a character vector for `c(2, 6, "3")` in the
challenge above is called **type coercion**.

When R encounters a *mix of data types* (here, numbers and characters)
to be combined into a single vector, it forces them all to be the same
type. It "must" do this because a vector can consist of only a single
data type.

Type coercion can be the source of many surprises, and is one reason you
need to be aware of the basic R data types and how R's behaviour around
them.

::: callout-note
### Manual Type Conversion

Luckily, you are not just at the mercy of whatever R decides to do
automatically, but can convert vectors using the `as.` group of
functions:

```{r}
as.integer(c("0", "2"))

as.character(c(0, 2))
```

As you may have guessed, though, not all type conversions are possible
--- for example:

```{r}
as.double("kiwi")
```
:::

### Missing values (`NA`)

R has a concept of **missing data**, which is important in statistical
computing, as not all information/measurements are always available for
each sample.

In R, missing values are coded as `NA` (like `TRUE`/`FALSE`, this is not
a character string so it is not quoted):

```{r}
# This vector will contain one missing value
vector_NA <- c(1, 3, NA, 7)
vector_NA
```

Notably, many functions operating on vectors will return `NA` if **any**
element in the vector is `NA`:

```{r}
sum(vector_NA)
```

You can get around this is by setting `na.rm = TRUE` in such functions,
for example:

```{r}
sum(vector_NA, na.rm = TRUE)
```

### Factors

**Categorical** data, like treatments in an experiment, can be stored as
"factors" in R. Factors are useful for statistical analyses and for
plotting, e.g. because they allow you to specify a custom order.

```{r}
diet_vec <- c("high", "medium", "low", "low", "medium")
diet_vec
factor(diet_vec)
```

In the example above, we turned a character vector into a factor. Its
"levels" (low, medium, high) are sorted alphabetically by default, but
we can manually specify an order that makes more sense:

```{r}
diet_fct <- factor(diet_vec, levels = c("low", "medium", "high"))
diet_fct
```

This ordering would be automatically respected in plots and statistical
analyses.

::: {.callout-warning collapse="true"}
### Oddly, factors are technically not a data type *(Click to expand)*

For most intents and purposes, it makes sense to think of factors as
another data type, even though technically, they are a kind of data
structure build on the `integer` data type:

```{r}
typeof(diet_fct)
```
:::

## Recap

-   Vectors: A one-dimensional array of elements of the same type.
-   Mixing types in vectors cause automatic coercion.
-   Elements of vectors can be accessed using `[ ]`.
-   Data types determines how data is stored, processed, and
    interpreted. For instance, treatments (factors), gene expression
    (double), gene names (character).

<br>
