---
title: "Week 11 Exercises: R Basics"
pagetitle: "Week 11 Exercises"
date: 2025-11-03
author:
  - name: Menuka Bhandari
    #orcid: 0000-0002-3514-7462
    email: bhandari.72@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://mcic.osu.edu/bioinformatics
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
other-links:
  - text: "Week 11 overview"
    icon: book
    href: w11_overview.html
  - text: "w11a: Intro to R and RStudio"
    icon: bookmarks
    href: w11a_r-basics.html
  - text: "w11b: R data structures and data types"
    icon: bookmarks
    href: w11b_data.html
  - text: "w11c: Tabular data in R"
    icon: bookmarks
    href: w11c_data-frames.html
editor_options: 
  chunk_output_type: console
---

--------------------------------------------------------------------------------

<hr style="height:1pt; visibility:hidden;" />

## Setting up

- At <https://ondemand.osc.edu>, start a new RStudio session
  ([Starting an RStudio Server session at OSC](../week11/w11a_r-basics.html#starting-an-rstudio-server-session-at-osc))

- Switch to your `week11` RStudio Project
  ([Working directory and RStudio Projects](../week11/w11a_r-basics.html#Working-directory-and-RStudio-Projects))
  
- Open a new R script in RStudio:
  `File` -> `New File` -> `R Script` and save the R script as 
  `w11_exercises.R` (inside `week11`).
   You will write your answers to the exercises below in this R script.

## Part A: Create variables and perfrom arithmetic operations

1. Assign the value `5` to the variable `week11`, and `10` to the variable `week12`.
   Print these variables to the console.

2. Add the values of `week11` and `week12` and assign the result to to a new variable `total_weeks`.
   Similarly, multiply the values of `week11` and `week12` and assign the result
   to a new variable `multiplied_weeks`.
   Print these variables to the console.

## Part B: Create vectors and name its elements

3. Create a vector `v1` that contains the numbers 1 through 5.
   Then, extract the third element of `v1`, printing it to the console.
   Likewise, create another variable `v2` with numbers 6 to 10.
   Check what data types `v1` and `v2` contain.

4. Change the name of the third element of `v1` to "Midweek" and print `v1`.
   Check the data type of `v1` after changing third element to `Midweek`.
   Try to change the `v1` to `numeric` and see what happens.

5. Create a vector `v3` that combines `v1` and `v2`.
   Extract the third, fourth and fifth element of `v3` and assign it to `v4`.
   Print `v4` to the console.

## Part C: Install and load new packages in R

6. Install the package `palmerpenguins` and load the package in R.
   We will use this package in `week13` for data visualization.

## Part D: Working with matrix and dataframes

7. Look for help for `matrix` function and create a matrix with numbers 1 to 10 with 2 rows and 5 columns
    and assign it to a new object called `my_matrix`.

8. Last week we worked with the `iris` data frame,
   which is one of the data sets that comes with R.
   For this exercise, we will use the `mtcars` data frame, which similarly comes with R.
   Start by storing `mtcars` in a new object called `my_cars`,
   then view the first 6 rows of `my_cars`.

9. Check the structure of the `my_cars` data frame.
   Structure of data shows the datatypes of different variables in the data frame.
   Compute the mean of the `mpg` variable of `my_cars` data frame.

10. Show the first 6 rows of the `my_cars` data frame. 
    Select columns `mpg`, `cyl`, and `hp` from the `my_cars` data frame and display the first 6 rows.

    ::: {.callout-tip appearance="simple"}
    Remember to load the _tidyverse_ package before using the `select()` function.
    :::

11. Write a single "pipeline" that manipulates `my_cars` by:
    - Selecting columns `mpg` (miles per gallon) and `cyl` (number of cylinders)
    - Only keeping rows with `mpg` greater than 10 and `cyl` equal to 6.
    - Renaming the `cyl` column to `cylinders`
    - Assigning the resulting data frame to an object called `my_cars_filtered`
  
12. Save the new_data frame as a CSV file called `my_cars_filtered.csv` in your working directory.
    In the RStudio "Files" pane,
    check that the file is there and click on it to view it in the editor pane.

## Solutions

### Part A: Create variables and perform arithmetic operations

<details><summary>**1.** Assign values to variables</summary>

```{r}
week11 <- 5
week11

week12 <- 10
week12
```
</details>

<details><summary>**2.** Arithmetic operations</summary>

```{r}
total_weeks <- week11 + week12
total_weeks

multiplied_weeks <- week11 * week12
multiplied_weeks
```
</details>

### Part B: Create and name a vector

<details><summary>**3.** Create vectors and extract elements</summary>

```{r}
v1 <- 1:5
v1[3]

v2 <- 6:10

str(v1)
str(v2)
```
</details>

<details><summary>**4.** Change vectors </summary>

**After replacing a numeric element with a string, the entire vector becomes a character vector**
**And when you try to convert it back to numeric, R will return `NA` for the non-numeric elements:**
   
```{r}
v1[3] <- "Midweek"
v1
str(v1)
as.numeric(v1)
```
</details>

<details><summary>**5.** Combine vectors </summary>

```{r}
v3 <- c(v1, v2)
v4 <- v3[3:5]
v4
```

### Part C: Install and load new packages in R

<details><summary>**6.** Install and load a package </summary>

```{r, eval=FALSE}
install.packages("palmerpenguins")
library(palmerpenguins)
```

### Part D: Working with matrix and dataframes

<details><summary>**7.** Create a matrix </summary>

```{r, eval=FALSE }
?matrix()
my_matrix <- matrix(1:10, nrow=2, ncol = 5)
my_matrix
```
</details>

<details><summary>**8.** Create object </summary>

```{r}
my_cars <- mtcars

head(my_cars)
```

</details>

<details><summary>**9.** Structure of data frame </summary>

```{r}
str(my_cars)

mean(my_cars$mpg)
```
</details>

<details><summary>**10.** Data manipulation </summary>

```{r}
head(my_cars)

library(tidyverse)

my_cars |>
  select(mpg, cyl, hp) |> 
  head()
```
</details>

<details><summary>**11.** Data manipulation and assign data to new object </summary>

```{r}
my_cars_filtered.csv <- my_cars |> 
  select(mpg, cyl) |>
  filter(mpg > 10, cyl == 6 ) |>
  rename(cylinders = cyl)

my_cars_filtered.csv
```
</details>

<details><summary>**13.** Save data in your working directory </summary>

```{r, eval=FALSE}
write_csv(my_cars_filtered.csv, "my_cars_filtered.csv")
```
