---
title: "Week 15 exercises: Additional gene expression analysis in R"
pagetitle: "Week 15 exercises"
date: 2026-11-30
date-modified: last-modified
author:
  - name: Menuka Bhandari
    #orcid: 0000-0002-3514-7462
    email: bhandari.72@osu.edu
    affiliation: 
      - name: CFAES Bioinformatics Core, Ohio State University
        url: https://go.osu.edu/bioinformatics-core
      - name: Dept. of Plant Pathology, Ohio State University
        url: https://plantpath.osu.edu
  - name: Jelmer Poelstra
    orcid: 0000-0002-3514-7462
    email: poelstra.1@osu.edu
editor_options: 
  chunk_output_type: console
other-links:
  - text: "Week 15 overview"
    icon: book
    href: w15_overview.html
  - text: "w15b: RNA-Seq count data analysis in R"
    icon: bookmarks
    href: w15b_deseq.html
language:
  title-block-published: "Week starts on"
  title-block-modified: "Last modified"
  other-links-title: "Also this week"
execute:
  eval: false
---

--------------------------------------------------------------------------------

<br>

## Setting up

```{r}
library(tidyverse)          # Misc. data manipulation and plotting
library(pheatmap)           # Heatmap plot
library(DESeq2)             # Differential expression analysis
```

```{r, echo=FALSE}
dds <- readRDS(here::here("data/dds.rds"))

```

## Heatmaps

Rather than plotting expression levels for many individual genes,
we can create "heatmap" plots to plot dozens (possibly even hundreds) of genes at once.

We will create heatmaps with the `pheatmap` function,
and let's make a heatmap for the top-25 most highly significant DEGs for our focal contrast.

Unlike with some of the functions we used before,
we unfortunately can't directly use our DESeq2 object,
but we have to extract and subset the count matrix,
and also pass the metadata to the heatmap function:

```{r}
dds_vst <- varianceStabilizingTransformation(dds)
```

```{r}
# We need a normalized count matrix, like for the PCA
# We can simply extract the matrix from the normalized dds object we created for the PCA
norm_mat <- assay(dds_vst)

# In the normalized count matrix, select only the genes of interest
# We'll reuse the 'top25_DE' vector that we created for the individual gene plots
norm_mat_sel <- norm_mat[match(top25_DE, rownames(norm_mat)), ]

# Sort the metadata
meta_sort <- meta |>
  arrange(treatment, time) |>
  select(treatment, time)
```

Now we can create the plot:

```{r}
pheatmap(
  norm_mat_sel,
  annotation_col = meta_sort,  # Add the metadata
  cluster_cols = FALSE,        # Don't cluster samples (=columns, cols)
  show_rownames = FALSE,       # Don't show gene names
  scale = "row",               # Perform z-scaling for each gene
  )
```

Notes on the code and plot above:

- The z-scaling with `scale =` will make sure we can compare genes with very different
  expression levels: after all, we're interested in relative expression levels across
  samples/sample groups.

- `pheatmap` will by default perform hierarchical clustering both at the sample (`col`)
  and gene (`row`) level, such that more similar samples and genes will appear closer to
  each other.
  Above, we turned clustering off for samples, since we want to keep them in their by-group order.

<hr style="height:1pt; visibility:hidden;" />

::: exercise
#### {{< fa user-edit >}} Bonus exercise: heatmaps

Make a heatmap with the top-25 **most-highly expressed** genes
(i.e., genes with the highest mean expression levels across all samples).

<details><summary>*Click for a hint: how to get that top-25*</summary>

```{r}
top25_hi <- names(sort(rowMeans(norm_mat), decreasing = TRUE)[1:25])
```

</details>

<details><summary>*Click for the solution*</summary>

```{r}
# In the normalized count matrix, select only the genes of interest
norm_mat_sel <- norm_mat[match(top25_hi, rownames(norm_mat)), ]

# Sort the metadata
meta_sort <- meta |>
  arrange(treatment, time) |>
  select(treatment, time)

# Create the heatmap
pheatmap(
  norm_mat_sel,
  annotation_col = meta_sort,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  scale = "row"
  )
```
</details>
:::
